White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-06-18 papers/wdmerger_II/paper.tex
Update paper II description of the integration

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI}.
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to append to them ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

@@

Given a set of nuclides and the reaction links between them, we now consider
how a burning step is performed in our software. The integration of the stiff,
implicit system is performed by \vode\ \citep{vode}, a copy of which is
provided with \castro. To \vode\ we provide an integration state containing
the molar abundances $Y_{n} = X_{n} / A_{n}$, where $X$ is the mass fraction
of the abundance in the zone and $A$ is the number of nucleons. The integration
of
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y})
\end{equation}
is determined using a right-hand-side provided by the nuclear burning network.
When it comes to the thermodynamical evolution, the internal energy $e$ of the zone
will change when the nuclear abundances evolve. During the integration we track
the energy release from the system using
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

In a hydrostatic burn, we keep $\rho$ and $T$ fixed throughout, and use
the energy released at the end to compute a final temperature that is
thermodynamically consistent with the new internal energy. By contrast,
in a self-heating burn, we allow the temperature to evolve in response
to the burning (see \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\left(\frac{\partial e}{\partial t} - \sum_n \frac{\partial e}{\partial X_n}\frac{\partial X_n}{\partial t}\right)
\end{equation}
Here $c_V$ is the specific heat at constant volume, and $(\partial e/\partial X_n)$
is the derivative of the internal energy with respect to the mass fraction of a given
species (with other thermodynamic variables held constant). Both quantities are provided by the
equation of state. During this burn, we can keep these thermodynamic
derivatives constant using their value at the beginning of the run, or at each step we
can choose to re-evaluate the equation of state using the latest value of $(\rho, T)$.
The latter is more expensive but also more accurate, and we use it in this paper.
A third option presented by \citet{raskin:2010} is a so-called ``hybrid'' mode.
In this mode, by default we do a hydrostatic burn. If that burn fails, or if the net
energy change is negative, we do the burn again in self-heating mode. All three options
are implemented in our burner software.

@@

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of all the
burning timesteps across all zones. To do this, we need a method for determining 
$\dot{e}$ and $\dot{X_k}$. A typical choice in the literature has been to set, for example,
\begin{equation}
  \dot{e} = \frac{e^{n} - e^{n-1}}{\Delta t^{n-1}}, \label{eq:burning_mode_4}
\end{equation}
where is $e^n$ is the internal energy at the start of the current timestep and
$e^{n-1}$ is the internal energy at the start of the previous timestep. 
The obvious analogue is used for constructing the species rate of change.
However, there are alternative methods of constructing this derivative estimate, 
and we have found that these different methods have measurable consequences.
We define four separate methods for calculating the time derivative, with 
the above being method 4. Method 3 is similar to method 4 but replaces the
denominator in Equation \ref{eq:burning_mode_4} with the change in 
internal energy over the last timestep only from the nuclear reactions.
Method 2 is the same as method 3 but we only use the change in internal 
energy from the most recent nuclear burning step, that is, the second-half
of the Strang-split burning from the last timestep (the denominator 
then becomes $\Delta t / 2$). In method 1, the most accurate option and 
the current default in \castro, we evaluate the right-hand-side of the 
burning network given the current state to explicitly obtain the 
instantaneous value of $\dot{e}$ and $\dot{X_k}$.

@@

With the timestep limited the way we advocate in this paper, 
the timesteps are generally short enough so that the errors 
due to splitting are small. Other approaches to the coupling 
between reactions and hydrodynamics have been proposed in the 
broader literature, especially iterative methods such as 
deferred corrections that allow each of these operators to 
feel the lagged effects of the other operators. For example,
in the context of low Mach number flows, \cite{nonaka:2012} have
used the method of spectral deferred corrections \cite{SDC} to
couple their advection-diffusion-reaction equation set. In our
context this would involve using the advection as a source term
for the burning network, and symmetrically including a source
term to the hydrodynamics corresponding to the energy release
from the burning. We are presently investigating such a method,
and it may form the basis of further work on this subject.

Now we return to a point we hinted at above. The timestep
will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion
$\Delta t \leq f_{X_k} \tau_{X_k}$ when the estimates for
$\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions
over the timestep. However, this can assumption can fail
quite radically during periods of runaway burning when
the rate of change of these quantities is highly nonlinear.
We do not want to neglect the errors caused by this approximation
because they may build up over an extended period of nonlinear
evolution and perhaps substantially change the final results.
To this end, we have implemented a timestep retry option in
\castro. The retry option works by first saving a copy of the
full state, and then doing the advance. We then explicitly
check whether the timestep criteria were violated during the advance,
using the time-centered energy and mass fractions and a
time-centered differencing of the old and new states to
calculate the evolution timescale. If so, we reject this
advance and restore the original state data. We then perform
a subcycled update, with the maximum timestep chosen so that
the criterion is satisfied in each subcycle. Note that this
occurs entirely on a single level and is separate from the
subcycling that is used by \boxlib\ to advance the different
AMR levels in time; the full AMR algorithm does not ever
know that the retry subcycling took place. This can help
substantially to mollify the problem described above, where
the time derivative estimation methods proposed were not
sufficient to obtain the desired limiting. Thus we strongly
recommend that hydrodynamics codes implement such an approach,
which ensures that the coupling actually does stay at the
level requested. The most significant drawback is that saving
a copy of the old state effectively doubles the memory footprint
of the code, so this option should be disabled when not needed.
However, it may be crucial in determining the evolution of
systems such as white dwarf collisions, where even a short
period of incorrect advances can significantly pollute the
final answer, and the additional memory expense is justified
by obtaining a more accurate solution.

@@

While this approach may add several AMR levels 
to a simulation, we agree with \citet{kushnir:2013} that 
solving this numerical instability is crucial to avoiding
early, unphysical detonations, and so it is unavoidable
if a correct evaluation of the burning phase is desired.
A simulation that does not solve this problem in this way
(or some analogous manner) will not obtain the correct amount 
of burning, and will not converge properly with resolution. 
Fortunately, the regions where this criterion is not satisfied 
are fairly localized on the grid, so the amount of additional 
computational work is not insurmountable for a single simulation.
However, this level of accuracy may make a parameter study of
the problem unfeasible, given current computational resources.

>>>>

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI}.
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to couple them to ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

@@

Given a set of nuclides and the reaction links between them, we now consider
how a burning step is performed in our software. 
provided with \castro. To \vode\ we provide an integration state containing
the molar abundances $Y_{n} = X_{n} / A_{n}$, where $X$ is the mass fraction
of the abundance in the zone and $A$ is the number of nucleons. The integration
of
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y})
\end{equation}
is determined using a right-hand-side provided by the nuclear burning network.
When it comes to the thermodynamical evolution, the internal energy $e$ of the zone
will change when the nuclear abundances evolve. During the integration we track
the energy release from the system using
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

In a hydrostatic burn, we keep $\rho$ and $T$ fixed throughout, and use
the energy released at the end to compute a final temperature that is
thermodynamically consistent with the new internal energy. By contrast,
in a self-heating burn, we allow the temperature to evolve in response
to the burning (see\footnote{In the cited paper, a term based on the
thermodynamic chemical potential was included; we now believe
that it is incorrect to include such a term in the burn, since it
automatically sums to zero analytically.} \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\frac{\partial e}{\partial t}
\end{equation}
(Although $T$ evolves during the burn so that the integration is physically
accurate, as for the hydrostatic method we discard the final value
for $T$ at the end of the burn and recompute a temperature for the zone that is
consistent with its new internal energy.) Here $c_V$ is the specific heat at
constant volume, which is provided by the equation of state.  During this burn,
we can keep $c_V$ constant using its initial value, or at each step we
can choose to re-evaluate the equation of state using the latest value of $(\rho, T)$.
The latter is more expensive but also more accurate, and we use it in this paper.
In practice we find that the cost is negligible in comparison to the more expensive
parts of the calculation, and it can significantly speed up convergence near NSE.
A third option presented by \citet{raskin:2010} is a so-called ``hybrid'' mode.
In this mode, by default we do a hydrostatic burn. If that burn fails, or if the net
energy change is negative, we do the burn again in self-heating mode. All three options
are implemented in our burner software. The simulations shown in this work all
use the self-heating mode unless otherwise specified.

In our \microphysics\ repository we provide several software options for
solving a set of coupled stiff ODEs. For this work we use an implementation of
the well known variable-order Richardson extrapolation method presented by Stoer and
Bulirsch \cite{stoer:1980}, that is similar to the integrator which ships with
the original versions of the networks mentioned above. Previous work using the \boxlib\
codes typically used the classic \vode\ integrator \cite{vode}. We do maintain a version
of the software that is compatible with our software interfaces in the \microphysics\
repository, but we have largely shifted to integrators which are written in modern
Fortran as a consequence of our efforts to run our codes on GPUs. The Stoer and Bulirsch
integrator we provide satisfies this criterion; we also provide a rewrite of the VODE
BDF algorithm that uses modern Fortran features. The Stoer and Bulirsch algorithm
relies on a uniform relative error tolerance for all of the ODEs in the system, which
we set at $10^{-6}$ for all simulations described here.

@@

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of all the
burning timesteps across all zones. To do this, we need a method for determining 
$\dot{e}$ and $\dot{X_k}$. A typical choice in the literature has been to set, for example,
\begin{equation}
  \dot{e} = \frac{e^{n} - e^{n-1}}{\Delta t^{n-1}}, \label{eq:burning_mode_4}
\end{equation}
where is $e^n$ is the internal energy at the start of the current timestep and
$e^{n-1}$ is the internal energy at the start of the previous timestep. 
The obvious analogue is used for constructing the species rate of change.
However, there are alternative methods of constructing this derivative estimate, 
and we have found that these different methods have measurable consequences.
We define four separate methods for calculating the time derivative, with 
the above being method 4. Method 3 is similar to method 4 but replaces the
denominator in \autoref{eq:burning_mode_4} with the change in 
internal energy over the last timestep only from the nuclear reactions.
Method 2 is the same as method 3 but we only use the change in internal 
energy from the most recent nuclear burning step, that is, the second-half
of the Strang-split burning from the last timestep (the denominator 
then becomes $\Delta t / 2$). In method 1, the most accurate option and 
the current default in \castro, we evaluate the right-hand-side of the 
burning network given the current state to explicitly obtain the 
instantaneous value of $\dot{e}$ and $\dot{X_k}$.

@@

With the timestep limited the way we advocate in this paper, 
the timesteps are generally short enough so that the errors 
due to splitting are small. Other approaches to the coupling 
between reactions and hydrodynamics have been proposed in the 
broader literature, especially iterative methods such as 
deferred corrections that allow each of these operators to 
feel the lagged effects of the other operators. For example,
in the context of low Mach number flows, \cite{nonaka:2012} have
used the method of spectral deferred corrections \citep{SDC} to
couple their advection-diffusion-reaction equation set. In our
context this would involve using the advection as a source term
for the burning network, and symmetrically including a source
term to the hydrodynamics corresponding to the energy release
from the burning. We are presently investigating such a method,
and it may form the basis of further work on this subject.

Now we return to a point we hinted at above. The timestep
will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion
$\Delta t \leq f_{X_k} \tau_{X_k}$ when the estimates for
$\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions
over the timestep. However, this can assumption can fail
during periods of runaway burning when the rate of change
of these quantities is highly nonlinear. We may not want
to neglect the errors caused by this approximation
because they may build up over an extended period of nonlinear
evolution and perhaps substantially change the final results.
To this end, we have implemented a timestep retry option in
\castro, which re-computes an advance if it violated the
stability criteria as judged from the end of the timestep.
However we have found that the benefits for this problem are
small and thus we do not use it for the simulations here.

@@

While this approach may add several AMR levels 
to a simulation, we agree with \citet{kushnir:2013} that 
solving this numerical instability is crucial to avoiding
early, unphysical detonations, and so it is unavoidable
if a correct evaluation of the burning phase is desired.
A simulation that does not solve this problem in this way
(or some analogous manner) will not obtain the correct amount 
of burning, and will not converge properly with resolution. 
Fortunately, the regions where this criterion is not satisfied 
are fairly localized on the grid, so the amount of additional 
computational work is not insurmountable for a single simulation.
However, this level of accuracy may make a parameter study of
the problem challenging given current computational resources.
When resources are limited, one can still use this method and set a
limit on the number of levels of refinement possible in the simulation,
so that the refinement will still go in the right places but may not
refine all of the way to the regime of guaranteed stability.
